Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-24 paper.tex
responded to a question of KC
revise intro

\begin{itemize}
\item threesomes \citep{siek2010threesomes,garcia2013calculating},
\item supercoercions \citep{garcia2013calculating}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
Recall that in these systems, casts are compressed using a composition
operator.  Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: D and
UD~\citep{siek2009exploring}. We are interested in the D blame
tracking strategy because it comes with a more straightforward notion
safe cast compared to UD~\citep{siek2009exploring}, which is why D was
chosen Grift compiler. We are also intersted in UD because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The UD hypercoercions were inspired by the
supercoercions of \citet{garcia2013calculating} (hence the name) and
the D hypercoercions were inspired by the normal forms of
\citet{siek2012interpretations}.
%
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that eager
strategies are incompatible with $\eta$-equivalence of functions.

Of course, an alternative cast representation must be proved
correct. This paper presents steps toward a general framework for
proving equivalence of cast calculi and, in particular, proves that an
abstract machine using \lazyD{} hypercoercions is equivalent to an
abstract machine using standard \lazyD{}
casts~\citep{siek2009exploring}. We conjecture that the framework can
be generalized to \lazyUD{} and that it can be applied to coercions in
normal form and threesomes.

To summarize, the primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving the correctness of \lazyD{} cast
  represantations.
\item a formal proof that hypercoercions respect the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

In Section~\ref{sec:background} we review cast calculi and coercions.
We present hypercoercions in
Section~\ref{sec:hypercoercion-definition}.  We present a framework
for proving correctness of cast reprensentations in
Section~\ref{sec:framework} and use it to prove the correctness of
\lazyD{} hypercoercions in
Section~\ref{sec:hypercoercion-correctness}.

@@

Fig.~\ref{fig:bisim-SC-D} defines the bisimulation relation.  It is
the smallest congruence relation that also includes the pairs induced
by the rules in Fig.~\ref{fig:bisim-SC-D}.
%
The relation is mutually defined on continuations, written $\kappa
\approx \kappa$, and values, written $v \approx v : T$, where $T$ is
their type.
%
If $k \approx \kappa$, adding an identity cast on top of $k$ gives a related 
continuation.
%
\todo[inline]{The use of ``wrap'' obscures this explanation.
  Consider getting rid of wrap by inlining it. -JS
\\---\\
I agree that inlining wrap and ext can improve readability of $\kappa \approx 
\kappa$, but this would make latex different from Agda, is it alright?
-KC}
%
If $\kappa_1 \approx \kappa_2$, extending $\kappa_1$ and $\kappa_2$ with the 
same cast gives related results.
%
If two values of type $P$ are related, injecting them to $\TOOdyn$ gives 
related values.
%
% The following is part of being a congruence. -JS
%Unit values are of course related. 
%
To relate to a naked function in \ineffCEKD, the function in \effCEK{C} 
must use identities as its casts.
%
If functions are related, we can apply the same cast on both side.
%
Rules for products and sums are similar.

>>>>

\begin{itemize}
\item threesomes \citep{siek2010threesomes,garcia2013calculating},
\item supercoercions \citep{garcia2013calculating}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
%Recall that in these systems, casts are compressed using a composition
%operator.
These systems compress casts using a compose operator.
Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
proof assistants, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: D and
UD~\citep{siek2009exploring}. We are interested in the D blame
tracking strategy because it comes with a more straightforward notion of
safe cast compared to UD~\citep{siek2009exploring}, which is why D was
chosen in Grift compiler~\citep{kuhlenschmidt2018efficient}. We are also 
interested in UD because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The UD hypercoercions were inspired by the
supercoercions of \citet{garcia2013calculating} (hence the name) and
the D hypercoercions were inspired by the normal forms of
\citet{siek2012interpretations}.
%
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that eager
strategies are incompatible with $\eta$-equivalence of functions.

Of course, an alternative cast representation must be proved
correct. This paper presents steps toward a general framework for
proving equivalence of cast calculi and, in particular, proves that an
abstract machine using \lazyD{} hypercoercions is equivalent to an
abstract machine using standard \lazyD{}
casts~\citep{siek2009exploring}. We conjecture that the framework can
be generalized to \lazyUD{} and that it can be applied to coercions in
normal form, threesomes, and supercoercions.

To summarize, the primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving the correctness of \lazyD{} cast
  represantations.
\item a formal proof that \lazyD\ hypercoercions respect the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

In Section~\ref{sec:background} we review cast calculi and coercions in normal 
form. 
We present hypercoercions in
Section~\ref{sec:hypercoercion-definition}.  We present a framework
for proving correctness of \lazyD\ cast reprensentations in
Section~\ref{sec:framework} and use it to prove the correctness of
\lazyD{} hypercoercions in
Section~\ref{sec:hypercoercion-correctness}.

@@

Fig.~\ref{fig:bisim-SC-D} defines the bisimulation relation.  It is
the smallest congruence relation that also includes the pairs induced
by the rules in Fig.~\ref{fig:bisim-SC-D}.
%
The relation is mutually defined on continuations, written $\kappa
\approx \kappa$, and values, written $v \approx v : T$, where $T$ is
their type.
%
If $k \approx \kappa$, adding an identity cast on top of $k$ gives a related 
continuation.
%
\todo[inline]{The use of ``wrap'' obscures this explanation.
  Consider getting rid of wrap by inlining it. -JS
\\---\\
I agree that inlining wrap and ext can improve readability of $\kappa \approx 
\kappa$, but this would make latex different from Agda, is it alright?
-KC
\\---\\
Yes, that is alright. Inlining is really a trivial change.
-JS
}
%
If $\kappa_1 \approx \kappa_2$, extending $\kappa_1$ and $\kappa_2$ with the 
same cast gives related results.
%
If two values of type $P$ are related, injecting them to $\TOOdyn$ gives 
related values.
%
% The following is part of being a congruence. -JS
%Unit values are of course related. 
%
To relate to a naked function in \ineffCEKD, the function in \effCEK{C} 
must use identities as its casts.
%
If functions are related, we can apply the same cast on both side.
%
Rules for products and sums are similar.
